---
title: "Genomic Prediction Analysis - Stage I of II: get BLUPs"
author: "wolfemd"
date: "2019-9-16"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = F, tidy = T)
```

# Objective

**Two-stage** genomic prediction refers to the following procedure:

**Stage 1:** Fit a linear mixed model to the data *without* genomic data. Individuals (e.g. clones / accessions) are modeled as independent and identically distributed (*i.i.d.*) random effects. The BLUPs for this random effect represent the measurable total genetic values of each individual. All the experimental design variation, e.g. replication and blocking effects have been controlled for in the creation of our new response variable, the BLUPs from the gneotype random effect.

**Stage 2:** Using a modified version of the BLUPs from step 1 as the response variable, fit a genomic prediction model, which now has reduced size because the number of observations is now the same as the number of individuals.

**NOTE:** In the animal breeding literature **single-step** often refers to predictions that combine pedigree and marker information simultaneously. That *is not* our meaning here.

The code below represents Stage I.

# Set-up training datasets
```{r, eval=F}
rm(list=ls()); gc()
library(tidyverse);library(magrittr)
trials<-readRDS("data/NaCRRI_ExptDesignsDetected_91619.rds")
phenos<-trials %>% 
      unnest(TrialData) %>% 
      select(programName,locationName,studyYear,TrialType,studyName,
             CompleteBlocks,IncompleteBlocks,
             yearInLoc,trialInLocYr,repInTrial,blockInRep,observationUnitDbId,
             germplasmName,FullSampleName,
             Trait,Value,MaxNOHAV,NOHAV,PropHAV) %>% 
      mutate(GID=ifelse(!is.na(FullSampleName),FullSampleName,germplasmName),
             IncompleteBlocks=ifelse(IncompleteBlocks==TRUE,"Yes","No"),
             CompleteBlocks=ifelse(CompleteBlocks==TRUE,"Yes","No")) %>% 
  filter(!is.na(GID)) %>% 
  group_by(Trait) %>% 
  nest(.key = "TrainingData")
rm(trials); gc()
```



# Fit IID models, get BLUPs from asreml
```{r, eval=F}
#library(asremlPlus)
library(furrr); library(asreml); options(mc.cores=18); plan(multiprocess)
phenos %<>%
  mutate(asFixedFormula="Value ~ yearInLoc",
         asFixedFormula=ifelse(grepl("logRTNO",Trait) | grepl("logFYLD",Trait) | grepl("logTOPYLD",Trait),
                               paste0(asFixedFormula," + PropHAV"),asFixedFormula),
         atBlockModel=paste0("~idv(GID) + idv(trialInLocYr) + at(CompleteBlocks,'Yes'):repInTrial ",
                              "+ at(IncompleteBlocks,'Yes'):blockInRep"),
         stdRandModel=paste0("~idv(GID) + idv(trialInLocYr) + idv(repInTrial) + idv(blockInRep)")) %>% 
      gather(Model,asRandFormula,atBlockModel:stdRandModel)
#phenos$TrainingData[[1]] %>% filter(is.na(GID)) %>% as.data.frame


asModelsFit<-phenos %>%
  mutate(fitAS=future_pmap(.,function(asFixedFormula,asRandFormula,TrainingData,...){
    out<-asreml(as.formula(asFixedFormula),
                         random = as.formula(asRandFormula),
                         data = TrainingData, maxiter = 40,workspace=400e6) 
    ll<-summary(out,all=T)$loglik
    varcomp<-summary(out,all=T)$varcomp
    aic<- -2*ll + 2*nrow(varcomp)
    Vg<-varcomp["GID!GID.var","component"]
    Ve<-varcomp["R!variance","component"]
    H2=Vg/(Vg+Ve)
    blups<-summary(out,all=T)$coef.random %>% 
          as.data.frame %>% 
          rownames_to_column(var = "GID") %>% 
          select(GID,solution,`std error`) %>% 
          filter(grepl("GID",GID)) %>% 
          rename(BLUP=solution) %>% 
          mutate(GID=gsub("GID_","",GID),
                 PEV=`std error`^2,
                 REL=1-(PEV/Vg),
                 drgBLUP=BLUP/REL,
                 WT=(1-H2)/((0.1 + (1-REL)/REL)*H2))
    out<-tibble(loglik=ll,AIC=aic,Vg,Ve,H2,
                blups=list(blups),varcomp=list(varcomp))
    return(out) }))

asModelsFit %>% 
      unnest(fitAS,.drop = T) %>% 
      select(Trait,Model,AIC) %>% 
      spread(Model,AIC) %>% 
      mutate(AICdiff=atBlockModel-stdRandModel,
             Diff=ifelse(AICdiff>2,"stdRandModel",
                         ifelse(AICdiff< -2,"atBlockBest","Similar"))) %>% count(Diff)
```
Diff             n
  <chr>        <int>
1 Similar          6
2 stdRandModel     3


Unlike IITA and NRCRI, seems clear that std. model is either best or just as good, so should use it. 

```{r, eval=F}
asModelsFit %<>% 
  select(-TrainingData,-asFixedFormula,-asRandFormula) %>% 
      filter(Model=="stdRandModel") %>% 
  unnest(fitAS)
saveRDS(asModelsFit,file="data/nacrri_blupsForCrossVal_91619.rds")

```
## Fit model, detect outliers
```{r, eval=F}
rm(list=ls()); gc()
library(tidyverse);library(magrittr)
trials<-readRDS("data/NaCRRI_ExptDesignsDetected_91619.rds")
phenos<-trials %>% 
      unnest(TrialData) %>% 
      select(programName,locationName,studyYear,TrialType,studyName,
             CompleteBlocks,IncompleteBlocks,
             yearInLoc,trialInLocYr,repInTrial,blockInRep,observationUnitDbId,
             germplasmName,FullSampleName,
             Trait,Value,MaxNOHAV,NOHAV,PropHAV) %>% 
      mutate(GID=ifelse(!is.na(FullSampleName),FullSampleName,germplasmName),
             IncompleteBlocks=ifelse(IncompleteBlocks==TRUE,"Yes","No"),
             CompleteBlocks=ifelse(CompleteBlocks==TRUE,"Yes","No")) %>% 
  filter(!is.na(GID)) %>% 
  group_by(Trait) %>% 
  nest(.key = "TrainingData")
rm(trials); gc()
library(furrr); library(asreml); options(mc.cores=18); plan(multiprocess)
phenos %<>%
  mutate(asFixedFormula="Value ~ yearInLoc",
         asFixedFormula=ifelse(grepl("logRTNO",Trait) | grepl("logFYLD",Trait) | grepl("logTOPYLD",Trait),
                               paste0(asFixedFormula," + PropHAV"),asFixedFormula),
         atBlockModel=paste0("~idv(GID) + idv(trialInLocYr) + at(CompleteBlocks,'Yes'):repInTrial ",
                              "+ at(IncompleteBlocks,'Yes'):blockInRep"),
         stdRandModel=paste0("~idv(GID) + idv(trialInLocYr) + idv(repInTrial) + idv(blockInRep)")) %>% 
      gather(Model,asRandFormula,atBlockModel:stdRandModel)

```

```{r, eval=F}
fitASmodelsWithOutlierDetect<-function(asFixedFormula,asRandFormula,TrainingData,...){
      out<-asreml(as.formula(asFixedFormula),
                  random = as.formula(asRandFormula),
                  data = TrainingData, 
                  maxiter = 40,workspace=400e6,aom=T)
      stdRes <- out$aom$R[,"stdCondRes"]
      nedf <- out$nedf 
      studRes <- stdRes / sqrt( (nedf - stdRes^2)/(nedf - 1) ) 
      outliers<-which(abs(studRes)>3.3)
      ll<-summary(out,all=T)$loglik
      varcomp<-summary(out,all=T)$varcomp
      vg<-varcomp["GID!GID.var","component"]
      ve<-varcomp["R!variance","component"]
      H2tmp<-vg/(vg+ve)
      blups<-summary(out,all=T)$coef.random %>% 
            as.data.frame %>% 
            rownames_to_column(var = "GID") %>% 
            dplyr::select(GID,solution,`std error`) %>% 
            filter(grepl("GID",GID)) %>% 
            rename(BLUP=solution) %>% 
            mutate(GID=gsub("GID_","",GID),
                   PEV=`std error`^2,
                   REL=1-(PEV/vg),
                   drgBLUP=BLUP/REL,
                   WT=(1-H2tmp)/((0.1 + (1-REL)/REL)*H2tmp))
      out<-list(loglik=ll,Vg=vg,Ve=ve,H2=H2tmp,
                blups=list(blups),
                varcomp=list(varcomp),
                Outliers=list(outliers))
      return(out) }
# fitVals <- fitted(out)
# library(txtplot)
# txtdensity(stdRes)
# txtplot(stdRes, fitVals)
```

```{r, eval=F}
# asFixedFormula<-phenos$asFixedFormula[[1]]
# asRandFormula<-phenos$asRandFormula[[1]]
# TrainingData<-phenos$TrainingData[[1]]
#rm(asFixedFormula,asRandFormula,TrainingData); gc()

library(asreml)
asModelsFit<-phenos %>%
      mutate(fitAS=future_pmap(.,fitASmodelsWithOutlierDetect),
             NoutR1=map_dbl(fitAS,~length(.$Outliers[[1]])),
             OutliersR1=map(fitAS,~.$Outliers[[1]]))
asModelsFit %>% dplyr::select(Trait,NoutR1)
asModelsFit %<>% 
      mutate(TrainingData=map2(TrainingData,fitAS,function(TrainingData,fitAS){
                   outliers2remove<-fitAS$Outliers[[1]]
                   out<-TrainingData[-outliers2remove,]
                   return(out) }))
asModelsFit %<>% 
      mutate(fitAS=future_pmap(.,fitASmodelsWithOutlierDetect),
             NoutR2=map_dbl(fitAS,~length(.$Outliers[[1]])),
             OutliersR2=map(fitAS,~.$Outliers[[1]]))
asModelsFit %>% dplyr::select(Trait,NoutR1,NoutR2)
asModelsFit %<>% 
      mutate(TrainingData=map2(TrainingData,fitAS,function(TrainingData,fitAS){
            outliers2remove<-fitAS$Outliers[[1]]
            out<-TrainingData[-outliers2remove,]
            return(out) }))
asModelsFit %<>% 
      mutate(fitAS=future_pmap(.,fitASmodelsWithOutlierDetect))

asModelsFit %<>%
      dplyr::select(-TrainingData,-asFixedFormula,-asRandFormula) %>% 
      mutate(fitAS=map(fitAS,as_tibble)) %>% 
      unnest(fitAS)

asModelsFit %<>% 
  filter(Model=="stdRandModel")


saveRDS(asModelsFit,file="data/nacrri_blupsForCrossVal_outliersRemoved_91619.rds")
```

# Next step

[Stage II: Cross-validation](CrossValidation.html)
